Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-11-26 _tex/manuscript.tex
Built site for gh-pages

\abstract{Purpose: According to current guidelines for elective nodal
irradiation of oropharyngeal squamous cell carcinoma (OPSCC) patients,
large parts of the contralateral lymphatic system are included in the
elective clinical target volume (CTV-N), even for lateralized tumors
without clinical lymph node involvement in the contralateral neck. In
this work, we present a probabilistic model for bilateral lymphatic
tumor progression in OPSCC to predict the personal risk for occult
disease in any lymph node level (LNL), given the patient's clinical
lymph node involvement, T-stage, and lateralization of the primary
tumor.

Methods: We extend a hidden markov model for lymphatic tumor
progression, which was previously developed for ipsilateral lymph node
involvement, to the contralateral neck. The model represents each of the
LNLs I, II, III, IV, V, and VII of both sides of the neck as a hidden
binary random variable with states healthy and involved. LNLs are
connected to the tumor and among each other via arcs that correspond to
spread probabilities. These spread probability rates are learned via
Markov chain Monte Carlo (MCMC) sampling from a dataset of 833 OPSCC
patients.

Results: The model is able to describe the data on lymph node
involvement well with a small number of interpretable parameters.
Midline extension of the primary tumor is the main risk factor for
contralateral involvement. In addition, the risk of occult metastases in
contralateral lymph node level increases with more advanced T-stage and
more severe ipsilateral involvement. The probability of involvement in
contralateral level III is very low if the upstream level II is
clinically negative. Similarly, the probability of involvement in
contralateral level IV is very low if the upstream level III is
clinically negative.

Conclusions: The model may guide personalized volume reduction of the
elective CTV-N. It suggests that the contralateral neck may be spared
from elective irradiation for lateralized tumors not crossing the
midline. In patients with primary tumors crossing the midline but
clinically negative contralateral neck, the contralateral elective
irradiation may be limited to level II.}

>>>>

\abstract{Purpose: Current guidelines for elective nodal irradiation in
oropharyngeal squamous cell carcinoma (OPSCC) recommend including large
portions of the contralateral lymphatic system in the clinical target
volume (CTV-N), even for lateralized tumors with no clinical lymph node
involvement in the contralateral neck. This study introduces a
probabilistic model of bilateral lymphatic tumor progression in OPSCC to
estimate personalized risks of occult disease in specific lymph node
levels (LNLs) based on clinical lymph node involvement, T-stage, and
tumor lateralization.

Methods: Building on a previously developed hidden Markov model for
ipsilateral lymphatic spread, we extend the approach to contralateral
neck involvement. The model represents LNLs I, II, III, IV, V, and VII
on both sides of the neck as binary hidden variables (healthy or
involved), connected via arcs representing spread probabilities. These
probabilities are learned using Markov chain Monte Carlo (MCMC) sampling
from a dataset of 833 OPSCC patients, enabling the model to reflect the
underlying lymphatic progression dynamics.

Results: The model accurately and precisely describes observed patterns
of lymph node involvement with a compact set of interpretable
parameters. Midline extension of the primary tumor is identified as the
primary risk factor for contralateral involvement, with advanced T-stage
and extensive ipsilateral involvement further increasing risk. The model
highlights key hierarchical dependencies: occult disease in
contralateral LNL III is highly unlikely if upstream LNL II is
clinically negative, and in contralateral LNL IV, occult disease is
exceedingly rare without LNL III involvement.

Conclusions: This model offers an interpretable, probabilistic framework
to inform personalized elective CTV-N volume reduction. For lateralized
tumors that do not cross the midline, it suggests the contralateral neck
may safely be excluded from elective irradiation. For tumors extending
across the midline but with a clinically negative contralateral neck,
elective irradiation could be limited to LNL II, reducing unnecessary
exposure while maintaining oncological safety.}

### 2024-11-26 manuscript.qmd
Built site for gh-pages

---
title: "Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer"
abstract: |
  Purpose: According to current guidelines for elective nodal irradiation of oropharyngeal squamous cell carcinoma (OPSCC) patients, large parts of the contralateral lymphatic system are included in the elective clinical target volume (CTV-N), even for lateralized tumors without clinical lymph node involvement in the contralateral neck. In this work, we present a probabilistic model for bilateral lymphatic tumor progression in OPSCC to predict the personal risk for occult disease in any lymph node level (LNL), given the patient's clinical lymph node involvement, T-stage, and lateralization of the primary tumor.

  Methods: We extend a hidden markov model for lymphatic tumor progression, which was previously developed for ipsilateral lymph node involvement, to the contralateral neck. The model represents each of the LNLs I, II, III, IV, V, and VII of both sides of the neck as a hidden binary random variable with states healthy and involved. LNLs are connected to the tumor and among each other via arcs that correspond to spread probabilities. These spread probability rates are learned via Markov chain Monte Carlo (MCMC) sampling from a dataset of 833 OPSCC patients.

  Results: The model is able to describe the data on lymph node involvement well with a small number of interpretable parameters. Midline extension of the primary tumor is the main risk factor for contralateral involvement. In addition, the risk of occult metastases in contralateral lymph node level increases with more advanced T-stage and more severe ipsilateral involvement. The probability of involvement in contralateral level III is very low if the upstream level II is clinically negative. Similarly, the probability of involvement in contralateral level IV is very low if the upstream level III is clinically negative.

  Conclusions: The model may guide personalized volume reduction of the elective CTV-N. It suggests that the contralateral neck may be spared from elective irradiation for lateralized tumors not crossing the midline. In patients with primary tumors crossing the midline but clinically negative contralateral neck, the contralateral elective irradiation may be limited to level II.
---

>>>>

---
title: "Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer"
abstract: |
  Purpose: Current guidelines for elective nodal irradiation in oropharyngeal squamous cell carcinoma (OPSCC) recommend including large portions of the contralateral lymphatic system in the clinical target volume (CTV-N), even for lateralized tumors with no clinical lymph node involvement in the contralateral neck. This study introduces a probabilistic model of bilateral lymphatic tumor progression in OPSCC to estimate personalized risks of occult disease in specific lymph node levels (LNLs) based on clinical lymph node involvement, T-stage, and tumor lateralization.

  Methods: Building on a previously developed hidden Markov model for ipsilateral lymphatic spread, we extend the approach to contralateral neck involvement. The model represents LNLs I, II, III, IV, V, and VII on both sides of the neck as binary hidden variables (healthy or involved), connected via arcs representing spread probabilities. These probabilities are learned using Markov chain Monte Carlo (MCMC) sampling from a dataset of 833 OPSCC patients, enabling the model to reflect the underlying lymphatic progression dynamics.

  Results: The model accurately and precisely describes observed patterns of lymph node involvement with a compact set of interpretable parameters. Midline extension of the primary tumor is identified as the primary risk factor for contralateral involvement, with advanced T-stage and extensive ipsilateral involvement further increasing risk. The model highlights key hierarchical dependencies: occult disease in contralateral LNL III is highly unlikely if upstream LNL II is clinically negative, and in contralateral LNL IV, occult disease is exceedingly rare without LNL III involvement.

  Conclusions: This model offers an interpretable, probabilistic framework to inform personalized elective CTV-N volume reduction. For lateralized tumors that do not cross the midline, it suggests the contralateral neck may safely be excluded from elective irradiation. For tumors extending across the midline but with a clinically negative contralateral neck, elective irradiation could be limited to LNL II, reducing unnecessary exposure while maintaining oncological safety.
---
